Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-08-01 paper.tex
Mention dropping anisotropic term for Phi
Convert todo notes to footnotes

% packages
\usepackage{amsmath}              % for \text, align, split ...
\usepackage{amssymb}              % for \gtrsim ...
\usepackage{braket}               % for \bra \ket
\usepackage{color}                % for colorful margin notes
\usepackage[mathscr]{euscript}    % for \mathscr
\usepackage{graphicx}             % for \includegraphics
\usepackage{hyperref}             % for \url and \href
\usepackage{listings}             % for typesetting code with \lstinline
\usepackage{tabularx}             % for tabularx

@@

\newcommand{\todo}[1]{\marginpar{\tiny{\textcolor{red}{#1}}}}
\renewcommand\todo[1]{} % if you wish to hide all todo notes

@@

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries, so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{ruff1996-leakage_part1,ross2003-leakage_part2,
  deat2013-leakage, pere2016-asl, radi2016-dynamical}
\todo{cite others}
capture the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\todo{discuss mc more clearly}
Monte Carlo methods like those
used in supernova simulations \cite{abdi2012-monte_carlo}
and stationary models of accretion disks \cite{rich2015-monte_carlo}
are an excellent tool but require large computational resources
because of their global treatment of the problem.

@@

The ray tracing framework
is conceptually simple because it solves the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann}) along characteristics,
reducing it to a one-dimensional ordinary differential equation.
It is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
with the history of that light-cone truncated at large optical depth.
\todo{state scaling $\mathscr{O}(n)$?}
And it easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.

@@

%\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
\todo{cite}
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions,
for example neutrino oscillations and neutrino-antineutrino annihilation,
\todo{cite}
depend sensitively on the angular distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\todo{justify these statements}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

@@

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\todo{clarify how momenta are decoupled}
When we're solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$, we have no
information about $f^{\nu_\sigma}(x^\alpha;p'_\beta)$, and
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$, which are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

@@

In Minkowski spacetime, and for a stationary observer
$u_\beta\rightarrow(-1,0,0,0)$, we would have $s=0$, $q=1$,
and $\varepsilon=-p_t$. In that case $\cos A$ may be identified
\todo{cite someone that uses $\mu$}
with the familiar forward direction cosine $\mu$.

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium.
By Eqn.~\ref{eqn:geodesic_x},
the affine parameter has dimension ${\rm length}\,{\rm energy}^{-1}$,
so the source term has dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
for the reactions
\begin{align}
  \nu_e + n            &\leftrightarrow e^- + p,\nonumber\\
  \bar{\nu}_e + p      &\leftrightarrow e^+ + n,\nonumber\\
  \nu_e + {}^AZ        &\leftrightarrow e^- + {}^A(Z+1),\nonumber\\
  \bar{\nu}_e + {}^AZ  &\leftrightarrow e^+ + {}^A(Z-1),\nonumber
\end{align}
with ${}^ZA$ representing a nucleus of mass number $A$ and charge $Z$.
In terms of the emissivity $j(\varepsilon)$ describing number of neutrinos
of energy $\varepsilon$ emitted per length,
and the absorption opacity $\chi_a(\varepsilon)$
describing the number absorbed per length,
the coefficients are
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the distribution function of neutrinos in radiative
equilibrium with the matter, i.e. the Fermi-Dirac distribution function
\begin{equation}
  \label{eqn:feq}
  f^{\rm eq}(\varepsilon) \equiv
  \left(1+e^{\varepsilon/(k_{\rm B} T) -\eta_\nu}\right)^{-1},
\end{equation}
with the neutrino chemical potentials dependent on the local density,
temperature, and composition of the fluid via the neutron, proton,
and electron chemical potentials:
$\eta_{\nu_e}=-\eta_{\bar{\nu}_e}=\eta_p-\eta_n+\eta_{e^-}$ and
$\eta_{\nu_x}=0$.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption \cite{burr2006-neutrino_opacities}.
By detailed balance of the absorption/emission reactions, we may alternatively
write the emissivity in terms of the equilibrium distribution function:
\begin{equation}
  \label{eqn:ae_emissivity_feq}
  \mathscr{E}_{\rm AE}(\varepsilon)
  = \mathscr{K}_{\rm AE}(\varepsilon) f^{\rm eq}(\varepsilon).
\end{equation}
See App.~\ref{ssec:sources_ae} for details.
\todo{identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
\todo{code these up}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \ell'^\gamma,
\end{align}
defining the specific energy density,
specific momentum density, and
specific radiation pressure tensor, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\todo{relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos a)\,db$.
\todo{consistent with Eqn.~\ref{eqn:def_direction}?}
We also make use of the number density and flux defined
\begin{align}
  \label{eqn:G}
  G(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta), \\
  \label{eqn:Ka}
  K^\mu(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu.
\end{align}

@@

We integrate Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} with adaptive step sizes,
using the 3rd order Runge-Kutta algorithm
which produces an error estimate by comparing the 3rd and 2nd order solutions.
After each step is taken, the errors for each of the nine variables of
Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} are compared.
If the maximum of these errors exceeds some absolute or relative threshold
\todo{specify ScaledAbsRel algorithm}
the step size is decreased and the step recomputed,
if it is below some threshold the next step size is increased.
% see OdeIntegrators::AdaptiveDense
% and OdeSteppers::RungeKutta3
% and OdeSteppers::DormandPrince5
% and OdeController
% and OdeControllers::ProportionalIntegral_Base::StepSizeToAttempt
% and OdeErrorMeasures::ScaledAbsRel
\todo{specify}

The volume data are components of the metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
\todo{state num independent components}
and the hydrodynamical data $W$, $u_i$, $\varrho$, $T$, and $Y_e$.
These fields are stored in spectral representation.
If they are computed via hydrodynamical simulation, they are saved to disk
at specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a quasi-stationary solution to the
\todo{clarify quasistationary timescale}
general relativistic hydrodynamics equations,
a single time is stored to disk, and no time interpolation is needed.

@@

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE,scat}+f_{\rm SE,scat}$
(labeled \emph{scat} since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE,noscat}$
(labeled \emph{noscat} since we ignore elastic scattering).
There is no simple relation between these forms
(i.e.\ $f_{\rm AE,scat} \neq f_{\rm AE,noscat}$)
because $f_{\rm AE}$ is integrated using the absorption plus scattering optical depth
in the first form and solely the absorption optical depth in the second form.
The differences between the two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the \emph{noscat} form of the equations has been the standard for
\todo{explain why; cite others}
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).
The opacity table is computed using \lstinline{NuLib} from cross-sections
neglecting weak magnetism
\todo{anything else?}
and is stored on a grid covering energy, density, temperature, and composition
ranges spanning $\varepsilon_i\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,6.3\times10^{15}]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,200]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.
Interpolation is performed linearly.

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t \frac{dt'}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\todo{could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

@@

As Fig.~\ref{fig:mfps-50km} shows, integrating the distribution functions
over the energies $\varepsilon\in[0,100]$~MeV
probes our numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\todo{discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

We sample the distribution function $f(\tilde{\varepsilon},\cos A,B)$
with ray tracing
over a uniform grid of 40 points in energy $\tilde{\varepsilon}\in[0,100]$
and 30 points in angle $\cos A\in[-1,1]$, holding fixed $B=\pi$.
Results are shown in Fig.~\ref{fig:avg_eps_doppler} for
the velocities $v=\{0,0.1,0.8\}$.
The ray tracing results are computed from total densities in each angular bin,
that is
\begin{equation}
  \label{eqn:avg_eps_per_cosA}
  \langle \tilde{\varepsilon} \rangle (\cos A) =
  \frac{\tilde{J}(\cos A)}{\tilde{G}(\cos A)},
\end{equation}
with the Eulerian energy density per angular bin given by
\begin{equation}
  \label{eqn:J_per_cosA}
  \tilde{J}(\cos A) =
  \frac{1}{(2\pi)^3} \int_0^{\infty} d\tilde{\varepsilon} \, \tilde{\varepsilon}^3
  \int_0^{2\pi} dB \, f(\tilde{\varepsilon},\cos A,B),
\end{equation}
and $\tilde{G}(\cos A)$ defined analogously.
\todo{add metric terms for these integrals}
Results are identical for all \emph{scat} and \emph{noscat} methods.
In Fig.~\ref{fig:avg_eps_doppler}
we see the common features of a red-shifted spectrum for receding fluid
($\cos A\ll0$),
a blue-shifted spectrum for approaching fluid
($\cos A\gg0$),
and a slightly red-shifted spectrum for fluid moving transverse to the
observer ($\cos A=0$).

@@

\subsection{Idealized Compact Star:
  Testing Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Optically thick star in Schwarszchild.
Total luminosity at infinite separation may be computed by
computing the radial energy flux at a finite separation from the star, $r$,
and using $p_t$ for the energy. We compute this asymptotic luminosity
at several radii, $r$, and show that the luminosity is independent of
the observer's radius.
\todo{compute $L$, put in figure}

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
To test our scattering treatments we calculate neutrino fields outside a
collapsed 15~${\rm M}_\odot$ star, 100~ms after core bounce,
comparing ray tracing fields to those from a Monte Carlo transport calculation.
Elastic scattering interior to the shock at $r\approx150$~km
significantly modifies the neutrinos' spectra,
\todo{demonstrate}
and the extended envelope outside the shock becomes a source of
higher-than-average-energy neutrinos.

>>>>

% packages
\usepackage{amsmath}              % for \text, align, split ...
\usepackage{amssymb}              % for \gtrsim ...
\usepackage{braket}               % for \bra \ket
\usepackage[mathscr]{euscript}    % for \mathscr
\usepackage{graphicx}             % for \includegraphics
\usepackage{hyperref}             % for \url and \href
\usepackage{listings}             % for typesetting code with \lstinline
\usepackage{tabularx}             % for tabularx

@@

@@

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries, so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{ruff1996-leakage_part1,ross2003-leakage_part2,
  deat2013-leakage, pere2016-asl, radi2016-dynamical}
\footnote{TODO: cite others}
capture the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\footnote{TODO: discuss mc more clearly}
Monte Carlo methods like those
used in supernova simulations \cite{abdi2012-monte_carlo}
and stationary models of accretion disks \cite{rich2015-monte_carlo}
are an excellent tool but require large computational resources
because of their global treatment of the problem.

@@

The ray tracing framework
is conceptually simple because it solves the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann}) along characteristics,
reducing it to a one-dimensional ordinary differential equation.
It is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
with the history of that light-cone truncated at large optical depth.
\footnote{TODO: state scaling $\mathscr{O}(n)$?}
And it easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.

@@

%\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
\footnote{TODO: cite}
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions,
for example neutrino oscillations and neutrino-antineutrino annihilation,
\footnote{TODO: cite}
depend sensitively on the angular distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\footnote{TODO: justify these statements}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

@@

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\footnote{TODO: clarify how momenta are decoupled}
When we're solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$, we have no
information about $f^{\nu_\sigma}(x^\alpha;p'_\beta)$, and
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$, which are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

@@

In Minkowski spacetime, and for a stationary observer
$u_\beta\rightarrow(-1,0,0,0)$, we would have $s=0$, $q=1$,
and $\varepsilon=-p_t$. In that case $\cos A$ may be identified
\footnote{TODO: cite someone that uses $\mu$}
with the familiar forward direction cosine $\mu$.

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\footnote{TODO: point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium.
By Eqn.~\ref{eqn:geodesic_x},
the affine parameter has dimension ${\rm length}\,{\rm energy}^{-1}$,
so the source term has dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
for the reactions
\begin{align}
  \nu_e + n            &\leftrightarrow e^- + p,\nonumber\\
  \bar{\nu}_e + p      &\leftrightarrow e^+ + n,\nonumber\\
  \nu_e + {}^AZ        &\leftrightarrow e^- + {}^A(Z+1),\nonumber\\
  \bar{\nu}_e + {}^AZ  &\leftrightarrow e^+ + {}^A(Z-1),\nonumber
\end{align}
with ${}^ZA$ representing a nucleus of mass number $A$ and charge $Z$.
In terms of the emissivity $j(\varepsilon)$ describing number of neutrinos
of energy $\varepsilon$ emitted per length,
and the absorption opacity $\chi_a(\varepsilon)$
describing the number absorbed per length,
the coefficients are
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the distribution function of neutrinos in radiative
equilibrium with the matter, i.e. the Fermi-Dirac distribution function
\begin{equation}
  \label{eqn:feq}
  f^{\rm eq}(\varepsilon) \equiv
  \left(1+e^{\varepsilon/(k_{\rm B} T) -\eta_\nu}\right)^{-1},
\end{equation}
with the neutrino chemical potentials dependent on the local density,
temperature, and composition of the fluid via the neutron, proton,
and electron chemical potentials:
$\eta_{\nu_e}=-\eta_{\bar{\nu}_e}=\eta_p-\eta_n+\eta_{e^-}$ and
$\eta_{\nu_x}=0$.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption \cite{burr2006-neutrino_opacities}.
By detailed balance of the absorption/emission reactions, we may alternatively
write the emissivity in terms of the equilibrium distribution function:
\begin{equation}
  \label{eqn:ae_emissivity_feq}
  \mathscr{E}_{\rm AE}(\varepsilon)
  = \mathscr{K}_{\rm AE}(\varepsilon) f^{\rm eq}(\varepsilon).
\end{equation}
See App.~\ref{ssec:sources_ae} for details.
\footnote{TODO: identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
\footnote{TODO: code these up}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \ell'^\gamma,
\end{align}
defining the specific energy density,
specific momentum density, and
specific radiation pressure tensor, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\footnote{TODO: relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos a)\,db$.
\footnote{TODO: consistent with Eqn.~\ref{eqn:def_direction}?}
We also make use of the number density and flux defined
\begin{align}
  \label{eqn:G}
  G(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta), \\
  \label{eqn:Ka}
  K^\mu(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu.
\end{align}

@@

We integrate Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} with adaptive step sizes,
using the 3rd order Runge-Kutta algorithm
which produces an error estimate by comparing the 3rd and 2nd order solutions.
After each step is taken, the errors for each of the nine variables of
Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} are compared.
If the maximum of these errors exceeds some absolute or relative threshold
\footnote{TODO: specify ScaledAbsRel algorithm}
the step size is decreased and the step recomputed,
if it is below some threshold the next step size is increased.
% see OdeIntegrators::AdaptiveDense
% and OdeSteppers::RungeKutta3
% and OdeSteppers::DormandPrince5
% and OdeController
% and OdeControllers::ProportionalIntegral_Base::StepSizeToAttempt
% and OdeErrorMeasures::ScaledAbsRel
\footnote{TODO: specify}

The volume data are components of the metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
\footnote{TODO: state num independent components}
and the hydrodynamical data $W$, $u_i$, $\varrho$, $T$, and $Y_e$.
These fields are stored in spectral representation.
If they are computed via hydrodynamical simulation, they are saved to disk
at specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a quasi-stationary solution to the
\footnote{TODO: clarify quasistationary timescale}
general relativistic hydrodynamics equations,
a single time is stored to disk, and no time interpolation is needed.

@@

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE,scat}+f_{\rm SE,scat}$
(labeled \emph{scat} since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE,noscat}$
(labeled \emph{noscat} since we ignore elastic scattering).
There is no simple relation between these forms
(i.e.\ $f_{\rm AE,scat} \neq f_{\rm AE,noscat}$)
because $f_{\rm AE}$ is integrated using the absorption plus scattering optical depth
in the first form and solely the absorption optical depth in the second form.
The differences between the two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the \emph{noscat} form of the equations has been the standard for
\footnote{TODO: explain why; cite others}
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).
The opacity table is computed using \lstinline{NuLib} from cross-sections
neglecting weak magnetism
\footnote{TODO: anything else?}
and is stored on a grid covering energy, density, temperature, and composition
ranges spanning $\varepsilon_i\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,6.3\times10^{15}]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,200]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.
Interpolation is performed linearly.

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t \frac{dt'}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\footnote{TODO: could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

@@

As Fig.~\ref{fig:mfps-50km} shows, integrating the distribution functions
over the energies $\varepsilon\in[0,100]$~MeV
probes our numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\footnote{TODO: discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

We sample the distribution function $f(\tilde{\varepsilon},\cos A,B)$
with ray tracing
over a uniform grid of 40 points in energy $\tilde{\varepsilon}\in[0,100]$
and 30 points in angle $\cos A\in[-1,1]$, holding fixed $B=\pi$.
Results are shown in Fig.~\ref{fig:avg_eps_doppler} for
the velocities $v=\{0,0.1,0.8\}$.
The ray tracing results are computed from total densities in each angular bin,
that is
\begin{equation}
  \label{eqn:avg_eps_per_cosA}
  \langle \tilde{\varepsilon} \rangle (\cos A) =
  \frac{\tilde{J}(\cos A)}{\tilde{G}(\cos A)},
\end{equation}
with the Eulerian energy density per angular bin given by
\begin{equation}
  \label{eqn:J_per_cosA}
  \tilde{J}(\cos A) =
  \frac{1}{(2\pi)^3} \int_0^{\infty} d\tilde{\varepsilon} \, \tilde{\varepsilon}^3
  \int_0^{2\pi} dB \, f(\tilde{\varepsilon},\cos A,B),
\end{equation}
and $\tilde{G}(\cos A)$ defined analogously.
\footnote{TODO: add metric terms for these integrals}
Results are identical for all \emph{scat} and \emph{noscat} methods.
In Fig.~\ref{fig:avg_eps_doppler}
we see the common features of a red-shifted spectrum for receding fluid
($\cos A\ll0$),
a blue-shifted spectrum for approaching fluid
($\cos A\gg0$),
and a slightly red-shifted spectrum for fluid moving transverse to the
observer ($\cos A=0$).

@@

\subsection{Idealized Compact Star:
  Testing Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Optically thick star in Schwarszchild.
Total luminosity at infinite separation may be computed by
computing the radial energy flux at a finite separation from the star, $r$,
and using $p_t$ for the energy. We compute this asymptotic luminosity
at several radii, $r$, and show that the luminosity is independent of
the observer's radius.
\footnote{TODO: compute $L$, put in figure}

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
To test our scattering treatments we calculate neutrino fields outside a
collapsed 15~${\rm M}_\odot$ star, 100~ms after core bounce,
comparing ray tracing fields to those from a Monte Carlo transport calculation.
Elastic scattering interior to the shock at $r\approx150$~km
significantly modifies the neutrinos' spectra,
\footnote{TODO: demonstrate}
and the extended envelope outside the shock becomes a source of
higher-than-average-energy neutrinos.
